# 

I want to investigate how the population prescribed drugs for anxiety, depression and psychosis vary geographically and how they have changed over a 10-year period at the Scottish national and local authority level.

1.	What are the rates of prescriptions at a national level?
2.	Are there differences in prescriptions between local authorities?
3.	If they have changed, how have they have changed over the ten years at a national level?
4.	If they have changed, how have they have changed over ten years between local authorities?

## Libraries

```{r "Import Libraries"}
library("tidyverse")
```

## Read in data files

Local and national data from the The Scottish Public Health Observatory (ScotPHO) Online Profiles Tool (2025) website. Available at: https://scotland.shinyapps.io/ScotPHO_profiles_tool/ (Accessed: 03 November 2025).

```{r}
all_data <- read_csv('./data/ScotPHO_Population prescribed drugs for anxiety depression and psychosis_allyears_national_and_council_extract_2025-11-03.csv')
```

## Tidy the data


```{r}
national_data <- all_data |> 
    filter(quintile == "Total" & area_type=="Scotland") |> 
    select(c("year", "value"))    

local_authority_data <- all_data |>
    filter(quintile == "Total" & area_type=="Council area") |> 
    select(c("year", "area_name", "value"))
```


```{r}
national_data |> 
    ggplot(aes(x=year, y=value)) +
    geom_path()
```


```{r}
local_authority_data |> 
    ggplot(aes(x=year, y=value, color=area_name)) +
    geom_path()
```