---
title: "Introduction to Data Health Science Report"
author: "Charlotte Skardon (52533106)"
output: pdf_document
---


# TOODS

[] Summary
[] data acquisition,
[] cleaning,
[] transformation,
[] analysis,
[] visualisation

I want to investigate how the population prescribed drugs for anxiety, depression and psychosis vary geographically 
and how they have changed over a 10-year period at the Scottish national and local authority level.

1. What are the rates of prescriptions at a national level?
2. Are there differences in prescriptions between local authorities?
3. If they have changed, how have they have changed over the ten years at a national level?
4. If they have changed, how have they have changed over ten years between local authorities?

## Summary


* Who is your target audience, and what is the key message you aim to convey through your visualisations? 
 
The target audience is government and public bodies, to help inform decision making at a local and national level. The key message to convey is how much the population prescribed drugs for anxiety, depression and psychosis has changed, how that varies regionally.

* What data did you use to produce your visualisation? What are the limitations of these data? 

The data used were from the Scottish Public Health Observatory (ScotPHO) [profiles tool](https://scotland.shinyapps.io/ScotPHO_profiles_tool/). The the population percentage prescribed drugs for anxiety, depression and psychosis indicator was selected for the Inequalities  'Council Area' and 'Scotland' datasets, covering the years from 2014 to 2023, and the 'total' value was used. 

*********The main limitations of the data is the granularity, it has the percentage of the 
population, but doesn't show the actual population, so it's difficult to guage how many people this affects.
For example, the Glasgow City is the most populous authority, and 22.5 - Orkney 19.6

* What are the strengths and limitations of the analytical and visualisation approach you adopted?

The line charts produced show the general trend

## Libraries

```{r "Import Libraries"}
library("tidyverse")
library("viridis")
library('plotly')
library('here')
library('ggrepel')
library('gghighlight')
```

## Data Acquisition - Read in data files

Local and national data from the The Scottish Public Health Observatory (ScotPHO) Online Profiles Tool (2025) website. Available at: https://scotland.shinyapps.io/ScotPHO_profiles_tool/ (Accessed: 03 November 2025).

```{r "Read in the ScotPHO data"}
df_all_data <- read_csv('./data/ScotPHO_Population prescribed drugs for anxiety depression and psychosis_allyears_national_and_council_extract_2025-11-03.csv')
```

## Cleaning and transforming the data

We're only actually interested in the `Total` data, so we'll filter on that, and run a quick summary to see what is in the data.

```{r "Check values of the data"}
df_all_data |> 
    # Filtering by the 'total' quintile, as that's what we're interested in
    filter(quintile == 'Total') |> 
    # Getting the count (`n`) and the min/max values for the `value` and `year` columns to see 
    # if anything is out of the expected range
    # Also checking for missing (`na`) values in the columns
    summarise(
        n = n(), 
        minValue = min(value, na.rm=TRUE), maxValue = max(value, na.rm=TRUE), missingValue = sum(is.na(value)),
        minYear = min(year), maxYear = max(year), missingYear = sum(is.na(year))
    )
```

* There are no missing values - so no need to filter where the value `!is.na(value)`
* `Year` is treated as a double, which is not correct

```{r "Mutate the type of the year to be an integer"}
df_all_data <- df_all_data |> 
    mutate(year = as.integer(year))
```

Now, we're creating two data frames, one for the national level data (where the `area_type` equals "Scotland") and one for the local authority level (where the `area_type` equals "Council area"). 

In both the `value` used is filtered by the `quintile` column, pulling out only the "Total" values.

The minimum columns needed are also selected, at the national level that is only the `year` and `value`, at the local authority level, the `area_name` is also selected.

In both cases, `value` is renamed to `percent_prescribed` to make it clearer when it it used later on what the `value` is.

```{r "Create dataframes for national and local authority sets"}
df_national_data <- df_all_data |> 
        # Filtering by 'Total' values and only for Scotland as a whole
        filter(quintile == "Total" & area_type=="Scotland") |> 
        # Only selecting the 'year' and 'value' columns as this is all that is needed.
        select(c("year", "value")) |> 
        # Rename value column to percent_prescribed for clarity
        rename(percent_prescribed = value)

df_local_authority_data <- df_all_data |>
        # Filtering by 'Total' values and where the 'area_type' is the Council Area to get Local Authority data
        filter(quintile == "Total" & area_type=="Council area") |> 
        # Selecting the 'year', 'value' and 'area_name' for the local authorities
        select(c("year", "area_name", "value")) |> 
        # Rename value column to percent_prescribed for clarity
        rename(percent_prescribed = value)
```

For one of my visualisations, I want to see the difference between 2014 and 2023 in the population percent prescribed 
drugs for anxiety, depression and psyhosis, so I filter by those two years, `pivot_wider` the data frame to allow me to 
add a `diff_percent` column to be rendered.

```{r "Work out the percent differences"}
df_local_authority_change <- df_local_authority_data |> 
    filter(year %in% c(2014, 2023)) |> 
    group_by(area_name) |> 
    pivot_wider(names_from = year, values_from = percent_prescribed) |> 
    mutate(diff_percent = `2023` - `2014`) 
```

I want to show the local authorities that have increased _more_ than the average from the National level, so I calculate that
here finding the values for `2014` and `2023` of the `percent_prescribed` and subtracting the `2014` value from the `2023` one.

```{r "Calculate the total difference"}
df_national_2014_2023 = df_national_data |> 
    # Filter years by 2014 and 2023
    filter(year %in% c(2014, 2023) ) |> 
    # Pivot those into columns to allow for subtraction
    pivot_wider(names_from = year, values_from = percent_prescribed)

# Set a variable to be the difference
national_diff = df_national_2014_2023$`2023` - df_national_2014_2023$`2014`
```

## Analysing the data

Running some basic stats for the `national` and `local_authority` data sets.

```{r}
df_national_data |> 
    summarise(
        # Count
        n = n(),
        # Mean 
        mean = mean(percent_prescribed), 
        # Standard Deviation
        sd = sd(percent_prescribed), 
        # Inter-quartile range
        iqr = IQR(percent_prescribed)
    )
df_local_authority_data |> 
    summarise(
        # Count
        n = n(), 
        # Mean
        mean = mean(percent_prescribed), 
        # Standard Deviation
        sd = sd(percent_prescribed), 
        # Inter-quartile range
        iqr = IQR(percent_prescribed)
    )
```

The biggest differences are the standard deviation (`sd`) and inter-quartile range (`iqr`) which shows there is a much wider range of values for the population prescribed drugs for anxiety, depression and psychosis on a local authority basis compared to national. The `mean` is the same which is what would be expected.

## Visualisation

First we render a line graph of the national level data, giving a single line.

```{r "National Data"}
df_national_data |> 
    # Plotting the `percent_prescibed` by `year`
    ggplot(aes(x=year, y=percent_prescribed)) +
    # Adding the points to the graph
    geom_point() +
    # Plot the line
    geom_path() + 
    # Setting the limits to the same as will be used for the 'local authority' data so the scales will match
    # Also adding a tick at every 0.5 percent to make it a bit clearer to read
    scale_y_continuous(limits=c(13.5,26), breaks = seq(13.5, 26, 0.5)) +
    # Setting the labels for the graph
    labs(
        title = "Population prescribed drugs for anxiety, depression and psychosis in Scotland",
        x = "Year",
        y = "Percent of Population"
    ) +
    # Using a minimal theme to make the graph less cluttered
    theme_minimal()
```

We then do the same but for the local authorities. The scale is kept the same to allow for comparison.

```{r "Local Authority Data"}
local_authority_plot <- df_local_authority_data |> 
    # Plotting the `percent_prescribed` by the `year` using the `area_name` for the colour
    ggplot(aes(x=year, y=percent_prescribed, colour=area_name)) +
    # Plot the lines
    geom_path() + 
    # From the min/max analysis above - I know these ranges are outside the min/max of the data
    scale_y_continuous(limits = c(13.5,26), breaks = seq(13.5, 26, 0.5)) +
    # Setting the labels
    labs(
        title = "Percent of Population prescribed drugs for anxiety,\ndepression and psychosis for each local authority in Scotland",
        x = "Year",
        y = "Percent",
        colour = "Local Authority"
    ) +
    # Using the Viridis colour scale for colour blindness safety
    scale_color_viridis(discrete = TRUE) +
    # I'm setting the panel background to be a different colour to attempt to make the 
    # lighter Viridis colours show up more
    theme(panel.background = element_rect(fill = '#f5f5f5ff', color = 'black'),
          panel.grid.major = element_line(color = 'lightgrey', linetype = 'dotted'),
          panel.grid.minor = element_line(color = 'lightgrey', size = .2))

# Rendering using Plotly to allow for an easier way to see which local authority is which
ggplotly(local_authority_plot)
```

We now render the change from the first year (`2014`) to the last (`2023`) for each of the local authorities and show the which of those authorities have 
changed the population prescribed drugs for anxiety, depression and psychosis more than the national average.

```{r}
df_local_authority_change |> 
    # Adding a field `isAbove` to be TRUE or FALSE if the change in percent is above (bigger) than the national average.
    mutate(isAbove = diff_percent > national_diff) |> 
    # Plot the percent difference by the area name. The fill is set to be by the `isAbove` field calculated above
    ggplot( aes(x=reorder(area_name, diff_percent), y=diff_percent, fill=factor(isAbove, c(TRUE, FALSE)))) + 
    # Rendering a bar chart to show the data
    geom_bar( stat="identity" ) +
    # Rendering it in a vertical form as there are too many areas to show in a horizontal way
    coord_flip() +
    # Adding the labels
    labs(
        title = "Difference between 2014 and 2023 in the percent of the population prescribed\ndrugs for anxiety, depression and psychosis by local authority in Scotland",
        x = "Local Authority Name",
        y = "Percent Difference",
        # Showing the legend, and what the national difference is (to 2 decimal places)
        fill = sprintf("Above the\nnational difference\n(%.2f %%)", national_diff)
    ) +    
    theme_minimal()
```
